---
title: "B705 - Week 4 Lecture Notes"
date: "January 30, 2024"
author: "Austin Allen"
format:
  html:
    self-contained: true
    page-layout: full
    title-block-banner: true
    toc: true
    toc-depth: 3
    toc-location: body
    number-sections: false
    html-math-method: katex
    code-fold: true
    code-summary: "Show the code"
    code-overflow: wrap
    code-copy: hover
    code-tools:
        source: false
        toggle: true
        caption: See code
execute: 
  warning: false
    
---



::: {.panel-tabset}

### Tuesday



**Hat Matrix**

* Whenever there's a tildae under a variable, that means it's a vector
* $H = X(X^TX)^{-1}X^T$
* The reason they call it the "hat" matrix is because it puts a hat on the Y:
  - $\hat{Y} = HY$
* H is symmetric and idempotent. Idempotent means that $HH = H$, i.e. you can multiply $H$ by itself and it won't change it
* The trace of an ideopotent matrix is equal to its rank. 

```{r}

A <- matrix(c(1,2,3,4,5,6), nrow = 2, byrow = TRUE)
B <- matrix(c(7,8,9,10,11,12), ncol = 2, byrow = TRUE)
C <- A %*% B
D <- B%*% A
C
D

```


**Interactions**


* If there is crossing over, there's definitely an interaction
* An interaction changes the slopes of the lines. For example, if we have a model like this:
  - $Y = \beta_0 + \beta_1 X_1 + \beta_2 X_2 + \beta_3 X_1 X_2$, 
  - we might have a model like the one below (assuming the interaction to be significant):

  ```{r}
library(tidyverse)
ggplot() +
  geom_segment(aes(x = 0, xend = 10, y = 0, yend = 10), col = "skyblue") +
  geom_segment(aes(x = 0, xend = 10, y = 10, yend = 0), col = "skyblue") +
  theme_bw()

  ```