---
title: "BIOSTAT 706 - Project 4"
output:
  html_document:
    theme: readable
---

```{r include = FALSE}
##### IMPORTANT: #####
# If you don't have any of these packages installed, you'll want to un-comment the code and install them

# install.packages("tidyverse")
# install.packages("here")
# install.packages("kableExtra")
# I'm pretty sure 'knitr' comes pre-loaded, but if you get errors, you should install `knitr` too

# Load libraries (make sure they're installed first)
library(tidyverse)
library(knitr)
library(here)
library(kableExtra)

# Set the global chunks for uniformity
knitr::opts_chunk$set(warning = FALSE, message = FALSE, fig.align = 'center', fig.width = 4, fig.height = 3)

# Set file paths for easy read-in
MIDUS_combined_reduced_file <- here::here("MIDUS_combined_reduced.csv")
new_variable_names_file <- here::here("new_variable_names.csv")
vars_to_keep_file <- here::here("vars_to_keep.csv")
age_score_data_file <- here::here("age_score_data.csv")

```


*Note:* Before knitting this file, make sure `MIDUS_combined_reduced.csv`, `new_variable_names.csv`, `vars_to_keep.csv` and `age_score_data.csv` are downloaded in the same folder as this file. 



## {.tabset .tabset-fade .tabset-pills}

### Assign Variables to Group

Everything in this section was used to create the .csv file called `new_variable_names.csv`. This code is not very important. 

```{r}
# Read in data and assign variable names to group for replacement (finished)
# dat <- read.csv("MIDUS_combined_reduced.csv")[569:1315,]
# column_names <- colnames(dat)
# 
# 

# kiki <- data.frame(current_variable_name = column_names[1:51], new_variable_name = rep(NA, 51))
# dorene <- data.frame(current_variable_name = column_names[52:102], new_variable_name = rep(NA, 51))
# qianyu <- data.frame(current_variable_name = column_names[103:153], new_variable_name = rep(NA, 51))
# tianyi <- data.frame(current_variable_name = column_names[154:204], new_variable_name = rep(NA, 51))
# austin <- data.frame(current_variable_name = column_names[205:254], new_variable_name = rep(NA, 50))
# 
# write.csv(austin, "austin.csv")
# write.csv(qianyu, "qianyu.csv")
# write.csv(tianyi, "tianyi.csv")
# write.csv(dorene, "dorene.csv")
# write.csv(kiki, "kiki.csv")

```

```{r}
# This code chunk is to perform basic wrangling on my own .csv file (finished)

# austin <- read.csv("austin.csv")
# 
# austin %>% 
#   mutate(description = new_variable_name,
#          new_variable_name = str_replace_all(str_to_lower(new_variable_name), " ", "_")) %>% 
#   write.csv("austin.csv")
```



```{r}
# Read in all returned variable replacement assignments (finished)
# kiki <- read.csv("variable_assignments/kiki.csv")[-1]
# dorene <- read.csv("variable_assignments/dorene.csv")[-1]
# qianyu <- read.csv("variable_assignments/qianyu.csv")[-1] %>% 
#   mutate(new_variable_name = ifelse(new_variable_name == "", current_variable_name, new_variable_name)) %>% 
#   rename(description = NOTEs)
# tianyi <- read.csv("variable_assignments/tianyi.csv")[-1]
# austin <- read.csv("variable_assignments/austin.csv")[-c(1,2)]

# Perform basic cleaning to get all names uniform (finished)
# new_variable_names <- rbind(kiki, dorene, qianyu, tianyi, austin) %>%
#    filter(new_variable_name != "") %>%
#    mutate(new_variable_name = str_to_lower(new_variable_name)) %>%
#    select(-description) %>% 
#   mutate(new_variable_name = str_replace_all(new_variable_name, " ", "_"))

# Save a copy
# write.csv(new_variable_names, "new_variable_names.csv")
```

### Assign new variable names to data

```{r}
# Load in data
dat <- read.csv(MIDUS_combined_reduced_file) 

dat <- dat %>% 
  filter(!is.na(MIDUS2), # Deselect individuals not in MIDUS2
         !is.na(MIDUS3)) # Deselect individuals not in MIDUS3 

# Load in new variable names (idetify those in M2 and M3)
new_variable_names <- read.csv(new_variable_names_file)[-1] 

new_variable_names <- new_variable_names %>% 
  mutate(new_variable_name = case_when(str_detect(new_variable_names$current_variable, '^B') ~ paste0("M3_", new_variable_name),
                                       str_detect(new_variable_names$current_variable, '^C') ~ paste0("M2_", new_variable_name),
                                       TRUE ~ new_variable_name))
                                       
# Rename variables
renamed_dat <- dat %>% 
  rename_with(~ new_variable_names$new_variable_name[match(., new_variable_names$current_variable_name)], everything())

# Store names in a character vector
new_names_vector <- colnames <- colnames(renamed_dat)

# how many are in M2 and M3?
m2_count <- sum(str_detect(new_names_vector, "M2"))
m3_count <- sum(str_detect(new_names_vector, "M3"))

# What columns are neither M2 nor M3? 
other_colnames <- new_names_vector[which(!str_detect(new_names_vector, "M[2,3]"))]


```


### Calculate Score


```{r}
# Load in the data
dat <- read.csv(MIDUS_combined_reduced_file) 

dat <-dat %>% 
  filter(!is.na(MIDUS2), # Deselect individuals not in MIDUS2
         !is.na(MIDUS3)) # Deselect individuals not in MIDUS3

# Load in the variables to keep
vars_to_keep <- read.csv(vars_to_keep_file)[,1]

# Select columns of interest
dat_selected <- dat[vars_to_keep]

# Values that are considered "missing"
missing_code <- c(7,8,9)

# Other conditions to be included (number shown in codebook for C4H1ZC)
included_other_conditions = c(2,3,4,5,6,7,9,10,13,14,15,16,17,18,19,20,21,22,24)


```


This next code chunk is the actual code that calculates the score. Note that this code is used to produce the file `age_score_data.csv`, which you should already have downloaded. 



```{r}
# Create the dataframe of interest
score_data <- dat %>% 
  mutate(heart_disease_3 = case_when(C4H1A == 1 ~ 1,
                                     C4H1A == 2 ~ .75,
                                     C4H1A == 3 ~ 0,
                                     C4H1A == 4 ~ .5,
                                     C4H1A == 5 ~ .25,
                                     TRUE ~ NA),
         high_blood_pressure_3 = case_when(C4H1B == 1 ~ 1,
                                     C4H1B == 2 ~ .75,
                                     C4H1B == 3 ~ 0,
                                     C4H1B == 4 ~ .5,
                                     C4H1B == 5 ~ .25,
                                     TRUE ~ NA),
         circulation_problems_3 = case_when(C4H1C == 1 ~ 1,
                                     C4H1C == 2 ~ .75,
                                     C4H1C == 3 ~ 0,
                                     C4H1C == 4 ~ .5,
                                     C4H1C == 5 ~ .25,
                                     TRUE ~ NA),         
         blood_clots_3 = case_when(C4H1D == 1 ~ 1,
                                     C4H1D == 2 ~ .75,
                                     C4H1D == 3 ~ 0,
                                     C4H1D == 4 ~ .5,
                                     C4H1D == 5 ~ .25,
                                     TRUE ~ NA),         
         heart_murmur_3 = case_when(C4H1E == 1 ~ 1,
                                     C4H1E == 2 ~ .75,
                                     C4H1E == 3 ~ 0,
                                     C4H1E == 4 ~ .5,
                                     C4H1E == 5 ~ .25,
                                     TRUE ~ NA),          
         stroke_3 = case_when(C4H1F == 1 ~ 1,
                                     C4H1F == 2 ~ .75,
                                     C4H1F == 3 ~ 0,
                                     C4H1F == 4 ~ .5,
                                     C4H1F == 5 ~ .25,
                                     TRUE ~ NA),          
         anemia_3 = case_when(C4H1G == 1 ~ 1,
                                     C4H1G == 2 ~ .75,
                                     C4H1G == 3 ~ 0,
                                     C4H1G == 4 ~ .5,
                                     C4H1G == 5 ~ .25,
                                     TRUE ~ NA),
         cholesterol_problems_3 = case_when(C4H1H == 1 ~ 1,
                                     C4H1H == 2 ~ .75,
                                     C4H1H == 3 ~ 0,
                                     C4H1H == 4 ~ .5,
                                     C4H1H == 5 ~ .25,
                                     TRUE ~ NA),          
         diabetes_3 = case_when(C4H1I == 1 ~ 1,
                                     C4H1I == 2 ~ .75,
                                     C4H1I == 3 ~ 0,
                                     C4H1I == 4 ~ .5,
                                     C4H1I == 5 ~ .25,
                                     TRUE ~ NA), 
         anemia_3 = case_when(C4H1J == 1 ~ 1,
                                     C4H1J == 2 ~ .75,
                                     C4H1J == 3 ~ 0,
                                     C4H1J == 4 ~ .5,
                                     C4H1J == 5 ~ .25,
                                     TRUE ~ NA),  
         copd_3 = case_when(C4H1K == 1 ~ 1,
                                     C4H1K == 2 ~ .75,
                                     C4H1K == 3 ~ 0,
                                     C4H1K == 4 ~ .5,
                                     C4H1K == 5 ~ .25,
                                     TRUE ~ NA), 
         tuberculosis_3 = case_when(C4H1L == 1 ~ 1,
                                     C4H1L == 2 ~ .75,
                                     C4H1L == 3 ~ 0,
                                     C4H1L == 4 ~ .5,
                                     C4H1L == 5 ~ .25,
                                     TRUE ~ NA), 
         thyroid_disease_3 = case_when(C4H1N == 1 ~ 1,
                                     C4H1N == 2 ~ .75,
                                     C4H1N == 3 ~ 0,
                                     C4H1N == 4 ~ .5,
                                     C4H1N == 5 ~ .25,
                                     TRUE ~ NA),
         peptic_ulcer_3 = case_when(C4H1O == 1 ~ 1,
                                     C4H1O == 2 ~ .75,
                                     C4H1O == 3 ~ 0,
                                     C4H1O == 4 ~ .5,
                                     C4H1O == 5 ~ .25,
                                     TRUE ~ NA),
         cancer_3 = case_when(C4H1P == 1 ~ 1,
                                     C4H1P == 2 ~ .75,
                                     C4H1P == 3 ~ 0,
                                     C4H1P == 4 ~ .5,
                                     C4H1P == 5 ~ .25,
                                     TRUE ~ NA),
         colon_polyp_3 = case_when(C4H1Q == 1 ~ 1,
                                     C4H1Q == 2 ~ .75,
                                     C4H1Q == 3 ~ 0,
                                     C4H1Q == 4 ~ .5,
                                     C4H1Q == 5 ~ .25,
                                     TRUE ~ NA),
         arthritis_3 = case_when(C4H1R == 1 ~ 1,
                                     C4H1R == 2 ~ .75,
                                     C4H1R == 3 ~ 0,
                                     C4H1R == 4 ~ .5,
                                     C4H1R == 5 ~ .25,
                                     TRUE ~ NA),
         glaucoma_3 = case_when(C4H1S == 1 ~ 1,
                                     C4H1S == 2 ~ .75,
                                     C4H1S == 3 ~ 0,
                                     C4H1S == 4 ~ .5,
                                     C4H1S == 5 ~ .25,
                                     TRUE ~ NA),
         cirrhosis_3 = case_when(C4H1T == 1 ~ 1,
                                     C4H1T == 2 ~ .75,
                                     C4H1T == 3 ~ 0,
                                     C4H1T == 4 ~ .5,
                                     C4H1T == 5 ~ .25,
                                     TRUE ~ NA),
         depression_3 = case_when(C4H1V == 1 ~ 1,
                                     C4H1V == 2 ~ .75,
                                     C4H1V == 3 ~ 0,
                                     C4H1V == 4 ~ .5,
                                     C4H1V == 5 ~ .25,
                                     TRUE ~ NA),
 heart_disease_2 = case_when(B4H1A == 1 & B4H1AD == 1 ~ 1, 
                             B4H1A == 1 & B4H1AD == 2 ~ .75,
                             B4H1A == 1 & B4H1AD %in% missing_code ~ .75,
                             B4H1A == 3 & B4H1AD == 1 ~ .5, 
                             B4H1A == 3 & B4H1AD == 2 ~ .25,
                             B4H1A == 3 & B4H1AD %in% missing_code ~ .25,
                             B4H1A == 2 ~ 0,
                             TRUE ~ NA
                             
                                     ),
         high_blood_pressure_2 = case_when(B4H1B == 1 & B4H1BD == 1 ~ 1, 
                             B4H1B == 1 & B4H1BD == 2 ~ .75,
                             B4H1B == 1 & B4H1BD %in% missing_code ~ .75,
                             B4H1B == 3 & B4H1BD == 1 ~ .5, 
                             B4H1B == 3 & B4H1BD == 2 ~ .25,
                             B4H1B == 3 & B4H1BD %in% missing_code ~ .25,
                             B4H1B == 2 ~ 0,
                             TRUE ~ NA),
         circulation_problems_2 = case_when(B4H1C == 1 & B4H1CD == 1 ~ 1, 
                             B4H1C == 1 & B4H1CD == 2 ~ .75,
                             B4H1C == 1 & B4H1CD %in% missing_code ~ .75,
                             B4H1C == 3 & B4H1CD == 1 ~ .5, 
                             B4H1C == 3 & B4H1CD == 2 ~ .25,
                             B4H1C == 3 & B4H1CD %in% missing_code ~ .25,
                             B4H1C == 2 ~ 0,
                             TRUE ~ NA),         
         blood_clots_2 = case_when(B4H1D == 1 & B4H1DD == 1 ~ 1, 
                             B4H1D == 1 & B4H1DD == 2 ~ .75,
                             B4H1D == 1 & B4H1DD %in% missing_code ~ .75,
                             B4H1D == 3 & B4H1DD == 1 ~ .5, 
                             B4H1D == 3 & B4H1DD == 2 ~ .25,
                             B4H1D == 3 & B4H1DD %in% missing_code ~ .25,
                             B4H1D == 2 ~ 0,
                             TRUE ~ NA),        
         heart_murmur_2 = case_when(B4H1E == 1 & B4H1ED == 1 ~ 1, 
                             B4H1E == 1 & B4H1ED == 2 ~ .75,
                             B4H1E == 1 & B4H1ED %in% missing_code ~ .75,
                             B4H1E == 3 & B4H1ED == 1 ~ .5, 
                             B4H1E == 3 & B4H1ED == 2 ~ .25,
                             B4H1E == 3 & B4H1ED %in% missing_code ~ .25,
                             B4H1E == 2 ~ 0,
                             TRUE ~ NA),         
         stroke_2 = case_when(B4H1F == 1 & B4H1FD == 1 ~ 1, 
                             B4H1F == 1 & B4H1FD == 2 ~ .75,
                             B4H1F == 1 & B4H1FD %in% missing_code ~ .75,
                             B4H1F == 3 & B4H1FD == 1 ~ .5, 
                             B4H1F == 3 & B4H1FD == 2 ~ .25,
                             B4H1F == 3 & B4H1FD %in% missing_code ~ .25,
                             B4H1F == 2 ~ 0,
                             TRUE ~ NA),         
         anemia_2 = case_when(B4H1G == 1 & B4H1GD == 1 ~ 1, 
                             B4H1G == 1 & B4H1GD == 2 ~ .75,
                             B4H1G == 1 & B4H1GD %in% missing_code ~ .75,
                             B4H1G == 3 & B4H1GD == 1 ~ .5, 
                             B4H1G == 3 & B4H1GD == 2 ~ .25,
                             B4H1G == 3 & B4H1GD %in% missing_code ~ .25,
                             B4H1G == 2 ~ 0,
                             TRUE ~ NA),
         cholesterol_problems_2 = case_when(B4H1H == 1 & B4H1HD == 1 ~ 1, 
                             B4H1H == 1 & B4H1HD == 2 ~ .75,
                             B4H1H == 1 & B4H1HD %in% missing_code ~ .75,
                             B4H1H == 3 & B4H1HD == 1 ~ .5, 
                             B4H1H == 3 & B4H1HD == 2 ~ .25,
                             B4H1H == 3 & B4H1HD %in% missing_code ~ .25,
                             B4H1H == 2 ~ 0,
                             TRUE ~ NA),          
         diabetes_2 = case_when(B4H1I == 1 & B4H1ID == 1 ~ 1, 
                             B4H1I == 1 & B4H1ID == 2 ~ .75,
                             B4H1I == 1 & B4H1ID %in% missing_code ~ .75,
                             B4H1I == 3 & B4H1ID == 1 ~ .5, 
                             B4H1I == 3 & B4H1ID == 2 ~ .25,
                             B4H1I == 3 & B4H1ID %in% missing_code ~ .25,
                             B4H1I == 2 ~ 0,
                             TRUE ~ NA),
         anemia_2 = case_when(B4H1J == 1 & B4H1JD == 1 ~ 1, 
                             B4H1J == 1 & B4H1JD == 2 ~ .75,
                             B4H1J == 1 & B4H1JD %in% missing_code ~ .75,
                             B4H1J == 3 & B4H1JD == 1 ~ .5, 
                             B4H1J == 3 & B4H1JD == 2 ~ .25,
                             B4H1J == 3 & B4H1JD %in% missing_code ~ .25,
                             B4H1J == 2 ~ 0,
                             TRUE ~ NA),  
         copd_2 = case_when(B4H1K == 1 & B4H1KD == 1 ~ 1, 
                             B4H1K == 1 & B4H1KD == 2 ~ .75,
                             B4H1K == 1 & B4H1KD %in% missing_code ~ .75,
                             B4H1K == 3 & B4H1KD == 1 ~ .5, 
                             B4H1K == 3 & B4H1KD == 2 ~ .25,
                             B4H1K == 3 & B4H1KD %in% missing_code ~ .25,
                             B4H1K == 2 ~ 0,
                             TRUE ~ NA),
         tuberculosis_2 = case_when(B4H1L == 1 & B4H1LD == 1 ~ 1, 
                             B4H1L == 1 & B4H1LD == 2 ~ .75,
                             B4H1L == 1 & B4H1LD %in% missing_code ~ .75,
                             B4H1L == 3 & B4H1LD == 1 ~ .5, 
                             B4H1L == 3 & B4H1LD == 2 ~ .25,
                             B4H1L == 3 & B4H1LD %in% missing_code ~ .25,
                             B4H1L == 2 ~ 0,
                             TRUE ~ NA),
         thyroid_disease_2 = case_when(B4H1N == 1 & B4H1ND == 1 ~ 1, 
                             B4H1N == 1 & B4H1ND == 2 ~ .75,
                             B4H1N == 1 & B4H1ND %in% missing_code ~ .75,
                             B4H1N == 3 & B4H1ND == 1 ~ .5, 
                             B4H1N == 3 & B4H1ND == 2 ~ .25,
                             B4H1N == 3 & B4H1ND %in% missing_code ~ .25,
                             B4H1N == 2 ~ 0,
                             TRUE ~ NA),
         peptic_ulcer_2 = case_when(B4H1O == 1 & B4H1OD == 1 ~ 1, 
                             B4H1O == 1 & B4H1OD == 2 ~ .75,
                             B4H1O == 1 & B4H1OD %in% missing_code ~ .75,
                             B4H1O == 3 & B4H1OD == 1 ~ .5, 
                             B4H1O == 3 & B4H1OD == 2 ~ .25,
                             B4H1O == 3 & B4H1OD %in% missing_code ~ .25,
                             B4H1O == 2 ~ 0,
                             TRUE ~ NA),
         cancer_2 = case_when(B4H1P == 1 & B4H1PD == 1 ~ 1, 
                             B4H1P == 1 & B4H1PD == 2 ~ .75,
                             B4H1P == 1 & B4H1PD %in% missing_code ~ .75,
                             B4H1P == 3 & B4H1PD == 1 ~ .5, 
                             B4H1P == 3 & B4H1PD == 2 ~ .25,
                             B4H1P == 3 & B4H1PD %in% missing_code ~ .25,
                             B4H1P == 2 ~ 0,
                             TRUE ~ NA),
         colon_polyp_2 = case_when(B4H1Q == 1 & B4H1QD == 1 ~ 1, 
                             B4H1Q == 1 & B4H1QD == 2 ~ .75,
                             B4H1Q == 1 & B4H1QD %in% missing_code ~ .75,
                             B4H1Q == 3 & B4H1QD == 1 ~ .5, 
                             B4H1Q == 3 & B4H1QD == 2 ~ .25,
                             B4H1Q == 3 & B4H1QD %in% missing_code ~ .25,
                             B4H1Q == 2 ~ 0,
                             TRUE ~ NA),
         arthritis_2 = case_when(B4H1R == 1 & B4H1RD == 1 ~ 1, 
                             B4H1R == 1 & B4H1RD == 2 ~ .75,
                             B4H1R == 1 & B4H1RD %in% missing_code ~ .75,
                             B4H1R == 3 & B4H1RD == 1 ~ .5, 
                             B4H1R == 3 & B4H1RD == 2 ~ .25,
                             B4H1R == 3 & B4H1RD %in% missing_code ~ .25,
                             B4H1R == 2 ~ 0,
                             TRUE ~ NA), 
          glaucoma_2 = case_when(B4H1S == 1 & B4H1SD == 1 ~ 1, 
                             B4H1S == 1 & B4H1SD == 2 ~ .75,
                             B4H1S == 1 & B4H1SD %in% missing_code ~ .75,
                             B4H1S == 3 & B4H1SD == 1 ~ .5, 
                             B4H1S == 3 & B4H1SD == 2 ~ .25,
                             B4H1S == 3 & B4H1SD %in% missing_code ~ .25,
                             B4H1S == 2 ~ 0,
                             TRUE ~ NA),
         cirrhosis_2 = case_when(B4H1T == 1 & B4H1TD == 1 ~ 1, 
                             B4H1T == 1 & B4H1TD == 2 ~ .75,
                             B4H1T == 1 & B4H1TD %in% missing_code ~ .75,
                             B4H1T == 3 & B4H1TD == 1 ~ .5, 
                             B4H1T == 3 & B4H1TD == 2 ~ .25,
                             B4H1T == 3 & B4H1TD %in% missing_code ~ .25,
                             B4H1T == 2 ~ 0,
                             TRUE ~ NA),
         depression_2 = case_when(B4H1V == 1 & B4H1VD == 1 ~ 1, 
                             B4H1V == 1 & B4H1VD == 2 ~ .75,
                             B4H1V == 1 & B4H1VD %in% missing_code ~ .75,
                             B4H1V == 3 & B4H1VD == 1 ~ .5, 
                             B4H1V == 3 & B4H1VD == 2 ~ .25,
                             B4H1V == 3 & B4H1VD %in% missing_code ~ .25,
                             B4H1V == 2 ~ 0,
                             TRUE ~ NA),
 sex = B1PRSEX,
 age_M2 = B1PAGE_M2,
 age_M3 = C1PRAGE)

score_data <- score_data[255:295] %>% 
  na.omit() %>% 
  mutate(score_M2 = rowSums(select(., matches("_2$"))),
         score_M3 = rowSums(select(., matches("_3$"))),
         mean_age_score = (score_M2 + score_M3)/2,
         mean_age = (age_M2 + age_M3)/2) 

  
```

```{r}
# You should already have this file downloaded
# write.csv(score_data, "age_score_data.csv")
```


### Table 


In the `Calculate Score` section, we wrote the file, `age_score_data.csv`. Let's print out a table of the data in this file. 

```{r}
# Load in score_data file
score_data <- read.csv(age_score_data_file)

knitr::kable(score_data, escape = FALSE) %>%
  kableExtra::kable_styling(full_width = T, scroll = TRUE) %>%
  mutate(
    # Set max-height for the table container (optional)
    attr("style") = paste(attr("style"), "max-height: 300px;", sep = "")
  )

```


### Plot

I don't remember exactly why I created this plot with sex as a factor, but here it is. 

```{r}
ggplot(score_data, aes(x = mean_age, y = mean_age_score, color = as.factor(sex))) +
  geom_point() +
  geom_smooth(method = "lm", se = FALSE, lty = 2, lwd = .75) +
  labs(x = "Mean Chronological Age", y = "Mean Aging Score", title = "How does the mean age score differ by sex?", col = "Sex") +
  scale_color_manual(values = c("skyblue", "pink")) +
  theme_minimal() + 
  theme(text = element_text(family = "serif"),
        plot.title = element_text(family = "serif", face = "bold", hjust = .5),
        panel.grid = element_blank())


```









